Minimum Regret Search for Single- and Multi-Task Optimization

### 2016-05-19 2016_ICML/cmrs.tex
Replacing "immediate regret" by "simple regret" and mentioning pure exploration setting

\begin{abstract}
We propose minimum regret search (MRS), a novel acquisition function for Bayesian
optimization. MRS bears similarities with information-theoretic approaches
such as entropy search (ES). However, while ES aims in each query at maximizing the
information gain with respect to the global maximum, MRS aims at minimizing the
expected immediate regret of its ultimate recommendation for the optimum. While empirically ES and MRS perform similar in most of the
cases, MRS produces fewer outliers with high immediate regret than ES. We provide empirical
results both for a synthetic single-task optimization problem as well as for a
simulated multi-task robotic control problem.
\end{abstract}

@@

Bayesian optimization \citep[BO,][]{shahriari_taking_2016} denotes a sequential,
model-based, global approach for optimizing black-box functions. It is particularly well-suited for problems
which are non-convex, do not necessarily provide derivatives, are expensive to evaluate (either
computationally, economically, or morally), and can potentially be noisy. Under
these conditions, there is typically no guarantee for finding the true optimum of the
function with a finite number of function evaluations. Instead, one often aims
at finding a solution which has small \emph{immediate regret} with regard to the true optimum, where 
``immediate regret'' denotes the difference of the true optimal function value 
and the function value of the ``solution'' selected by the algorithm after a finite number of function evaluations. BO
aims at finding such a solution of small immediate regret while  minimizing at the same
time the number of evaluations of the expensive target function. For this, BO
maintains a probabilistic surrogate model of the objective function, and a
myopic utility or acquisition function, which defines the ``usefulness'' of
performing an additional function evaluation at a certain input for learning about the optimum.

@@

A critical component for the performance of BO is the acquisition function,
which controls the exploratory behavior of the sequential search procedure.
Different kinds of acquisition functions have been proposed, ranging from
improvement-based acquisition functions over optimistic acquisition functions to
information- theoretic acquisition functions (see Section
\ref{Section:Background}). In the latter class, the group of entropy
search-based approaches
\cite{villemonteix_informational_2008, hennig_entropy_2012,hernandez-lobato_predictive_2014}, 
which aims at maximizing the information gain regarding
the true optimum, has achieved state-of-the-art performance on a number of
synthetic and real-world problems. However, performance is often reported as the
median over many runs, which bears the risk that the median masks ``outlier'' runs that
perform considerably worse than the rest. In fact, our results indicate that
the performance of sampling-based entropy search is not necessarily better
than traditional and cheaper acquisition functions according to the mean immediate regret.

In this work, we propose minimum regret search (MRS), a novel acquisition
function that explicitly aims at minimizing the expected immediate regret
(Section \ref{Section:MRS}). MRS performs well according to both the mean and
median performance on a synthetic problem (Section
\ref{Section:ResultsSingleTask}). Moreover, we discuss how MRS can be extended
to multi-task optimization problems (Section \ref{Section:CMRS}) and present
empirical results on a simulated robotic control problem (Section
\ref{Section:ResultsMultiTask}).

@@

While information-based policies for Bayesian optimization such as ES and PES
have performed very well empirically, their internal objective of minimizing
the uncertainty in the location of the optimizer $\mathbf{x}^\star$, i.e.,
minimizing the differential entropy of $p_\star$, is actually different (albeit
related) to the common external objective of minimizing the immediate regret of
$\mathbf{\tilde x}_N$, the recommendation of BO for $\mathbf{x}^\star$ after $N$
trials, where we define the immediate regret of $\mathbf{\tilde x}_N$ as
$R_f(\mathbf{\tilde  x}_N) = f(\mathbf{x}^\star) - f(\mathbf{\tilde  x}_N) = \max_{\mathbf{x}} f(\mathbf{x}) - f(\mathbf{\tilde x}_N)$.

Clearly, $\mathbf{x}^\star$ has zero and thus minimum immediate regret, but a query that
results in the maximal decrease in $H(\mathbf{x}^\star)$ is not necessarily the
one that also results in the maximal decrease in the expected immediate regret
as we will discuss in Section \ref{Section:MRS_Illustration}. In this section, we propose \emph{minimum
regret search} (MRS), which explicitly aims at minimizing the expected immediate
regret.

\subsection{Formulation}
Let $\mathcal{X} \subset \mathcal{R}^D$ be some bounded domain and $f: \mathcal{X} \mapsto
\mathbb{R}$ be a function. We are interested in finding the maximum
$\mathbf{x}^\star$ of $f$ on $\mathcal{X}$. Let there be a probability measure
$p(f)$ over the space of functions $f: \mathcal{X} \mapsto \mathbb{R}$, such as
a GP. Based on this $p(f)$, we would ultimately like to
select an $\mathbf{\tilde x}$ which has minimum immediate regret
$R_f(\mathbf{\tilde x})$. We define the expected immediate regret ER of
selecting parameter $\mathbf{x}$ under $p(f)$ as follows:
$$\text{ER}(p)(\mathbf{x}) = \mathbb{E}_{p(f)}[R_f(\mathbf{x})]
= \mathbb{E}_{p(f)}[\max_{\mathbf{x}} f(\mathbf{x}) - f(\mathbf{
x})]$$
The minimizer $\mathbf{\tilde x}$ of the expected immediate regret under fixed $p(f)$ can be determined efficiently in the case of a GP since it is identical to the maximizer of the GP's mean $\mu_N(\mathbf{x})$. However, in general, $p(f)$ depends on data and it is desirable to select the data such that $\text{ER}$ is also minimized with regard to the resulting $p(f)$.

@@

As choosing the optimal sequence $X^q_N$ at once is intractable, we follow the
common approach in BO and select $\mathbf{x}^q_n$
sequentially in a myopic way. However, as $\mathbf{\tilde x}_N$ itself depends
on $\mathcal{D}_N$ and is thus unknown for $n < N$, we have to use proxies for it based on the currently
available subset $\mathcal{D}_n \subset \mathcal{D}_N$.  One simple choice for a proxy is to use
the point which has minimal expected immediate regret under $p_n(f) =
p(f \vert \mathcal{D}_n)$, i.e.,
$\mathbf{\tilde x} = \arg\min_\mathbf{x} \text{ER}(p_n)(\mathbf{x})$.
%Based on this, we define the one-step
%(myopic) expected regret reduction with respect to a given $p(f)$ when
%evaluating the function at $\mathbf{x}^q$.
Let us denote the updated
probability measure on $f$ after performing a query at $\mathbf{x}^q$ and
observing the function value $y=f(\mathbf{x}^q) + \epsilon$ by
$p^{[\mathbf{x}^q, y]}_n = p(f\vert \mathcal{D}_n \cup \{\mathbf{x}^q, y\})$.
We define the acquisition function MRS$^{\text{point}}$ as the expected reduction of the minimum expected regret for a query at
$\mathbf{x}^q$, i.e.,
\begin{equation*}
\begin{split}
a_{\text{MRS}^{\text{point}}}(\mathbf{x}^q)
    & = \min_{\mathbf{\tilde x}}\text{ER}(p_n)(\mathbf{\tilde x}) \\
     &- \mathbb{E}_{y \vert p_n(f), \mathbf{x}^q}[\min_{\mathbf{\tilde x}}  \text{ER}(p^{[\mathbf{x}^q, y]}_n)(\mathbf{\tilde x})],
\end{split}
\end{equation*}
where the expectation is with
respect to $p_n(f)$'s predictive distribution at $\mathbf{x}^q$ and we drop the
implicit dependence on $p_n(f)$ in the notation of $a_{\text{MRS}^{\text{point}}}$.
The next query point $\mathbf{x}^q_{n+1}$ would thus be selected as the maximizer of
$a_{\text{MRS}^{\text{point}}}$.

@@

MRS can thus be seen as a more Bayesian treatment, where we marginalize our
uncertainty about $\mathbf{\tilde x}_N$, while MRS$^\text{points}$ is more akin
to a point-estimate since we use a single point (the minimizer of the expected immediate regret)
as proxy for $\mathbf{\tilde x}_N$.

@@

Figure \ref{fig:MRS_illustration} presents an illustration of different
acquisition functions on a simple one-dimensional target function. The left
graphic shows a hypothetical GP posterior (illustrated by its mean and standard
deviation) for length scale $l=0.75$, and the resulting probability of $\mathbf{x}$
being the optimum of $f$ denoted by $p^\star$.
Moreover, the expected immediate regret of selecting $\mathbf{x}$ denoted by
$\text{ER}(\mathbf{x})$ is shown. The minimum of $\text{ER}(\mathbf{x})$ and the maximum
of $p^\star$ are both at $\mathbf{\tilde x}=1.5$. The expected regret of $\mathbf{\tilde x}$
is approximately $\text{ER}(\mathbf{\tilde x}) = 0.07$. We plot $\mathbb{E}[\max(f(\mathbf{x}) -
f(\mathbf{\tilde x}), 0)]$ additionally to shed some light onto situations in which $
\mathbf{\tilde x} = 1.5$ would incur a significant regret: this quantity shows that most of the
expected regret of $\mathbf{\tilde x}$ stems from situations where the ``true''
optimum is located at $\mathbf{x} \geq 3.5$. This can be explained by the
observation that this area has high uncertainty and is at the same time largely
uncorrelated with $\mathbf{\tilde x} = 1.5$ because of the small length-scale of the GP.

@@

\begin{figure*}
\centering
\includegraphics[width=.9\textwidth]{empirical_comparison}
\includegraphics[width=.9\textwidth]{hist}
\caption{(Top) Median and mean immediate regret over $250$ repetitions for different acquisition functions. Shown is the immediate regret of the recommendation $\mathbf{\tilde x}_N$ after $N$ trials, i.e., the point which maximizes the GP posterior
mean. (Bottom) Histogram of the immediate regret after performing $N=100$ trials for different acquisition functions (note the log-scales).}
\label{fig:empirical_comparison}
\end{figure*}

@@

Figure \ref{fig:empirical_comparison} summarizes the results: according to the
median of the immediate regret (top left), ES, MRS, and MRS$^{point}$ perform
nearly identical, while EI is about an order of magnitude worse. GP-UCB performs
even worse initially but surpasses EI eventually\footnote{GP-UCB reaches 
the same level of mean and median immediate regret  as MRS after $N=200$ steps 
(in mean and median) with no significant difference  according to a Wilcoxon signed-rank test.}. 
PI performs the worst as it
exploits too aggressively. These results are roughly in correspondence with prior
results \cite{hennig_entropy_2012,hernandez-lobato_predictive_2014} on the same
task; note, however, that \citet{hernandez-lobato_predictive_2014} used a lower
noise level and thus, absolute values are not comparable. However, according to
the \emph{mean} immediate regret, the picture changes considerably (top right):
here, MRS, MRS$^{point}$, and EI perform roughly on par while ES is about an order of
magnitude worse. This can be explained by the distribution of the immediate
regrets (bottom): while the distributions are fairly non-normal for all
acquisition functions, there are considerably more runs with very high immediate
regret ($R_f(\mathbf{\tilde x}_N)>10^{-2}$) for ES (10) than for MRS (4) or EI (4).

We illustrate one such case where ES incurs high immediate regret in Figure
\ref{fig:es_analysis}. The same set of representer points has been used for ES and MRS.
While both ES and MRS assign a non-zero density $p^\star$
(representer points) to the area of the true optimum of the function
(bottom center), ES assigns high acquisition value only to areas with a high
density of $p^\star$ in order to further concentrate density in these areas.
Note that this is not due to discretization of $p^\star$ but because of ES'
objective, which is to learn about the precise location of the optimum, irrespective of
how much correlation their is between the representer points according to $p(f)$.
Thus, predictive entropy search \cite{hernandez-lobato_predictive_2014} would likely be affected by the
same deficiency. In contrast, MRS focuses first on areas which have not been
explored and have a non-zero $p^\star$, since those are areas with high expected
immediate regret (see Section \ref{Section:MRS_Illustration}). Accordingly, MRS is less likely to incur a high immediate regret. In summary, the MRS-based acquisition functions are the only
acquisition functions that perform well both according to the median and the
mean immediate regret; moreover, MRS performs slightly better than
MRS$^{point}$ and we will focus on MRS subsequently.

>>>>

\begin{abstract}
We propose minimum regret search (MRS), a novel acquisition function for Bayesian
optimization. MRS bears similarities with information-theoretic approaches
such as entropy search (ES). However, while ES aims in each query at maximizing the
information gain with respect to the global maximum, MRS aims at minimizing the
expected simple regret of its ultimate recommendation for the optimum. While empirically ES and MRS perform similar in most of the
cases, MRS produces fewer outliers with high simple regret than ES. We provide empirical
results both for a synthetic single-task optimization problem as well as for a
simulated multi-task robotic control problem.
\end{abstract}

@@

Bayesian optimization \citep[BO,][]{shahriari_taking_2016} denotes a sequential,
model-based, global approach for optimizing black-box functions. It is particularly well-suited for problems
which are non-convex, do not necessarily provide derivatives, are expensive to evaluate (either
computationally, economically, or morally), and can potentially be noisy. Under
these conditions, there is typically no guarantee for finding the true optimum of the
function with a finite number of function evaluations. Instead, one often aims
at finding a solution which has small \emph{simple regret} \cite{bubeck_pure_2009}) with regard to the true optimum, where 
simple regret denotes the difference of the true optimal function value 
and the function value of the ``solution'' selected by the algorithm after a finite number of function evaluations. BO
aims at finding such a solution of small simple regret while  minimizing at the same
time the number of evaluations of the expensive target function. For this, BO
maintains a probabilistic surrogate model of the objective function, and a
myopic utility or acquisition function, which defines the ``usefulness'' of
performing an additional function evaluation at a certain input for learning about the optimum.

@@

A critical component for the performance of BO is the acquisition function,
which controls the exploratory behavior of the sequential search procedure.
Different kinds of acquisition functions have been proposed, ranging from
improvement-based acquisition functions over optimistic acquisition functions to
information- theoretic acquisition functions (see Section
\ref{Section:Background}). In the latter class, the group of entropy
search-based approaches
\cite{villemonteix_informational_2008, hennig_entropy_2012,hernandez-lobato_predictive_2014}, 
which aims at maximizing the information gain regarding
the true optimum, has achieved state-of-the-art performance on a number of
synthetic and real-world problems. However, performance is often reported as the
median over many runs, which bears the risk that the median masks ``outlier'' runs that
perform considerably worse than the rest. In fact, our results indicate that
the performance of sampling-based entropy search is not necessarily better
than traditional and cheaper acquisition functions according to the mean simple regret.

In this work, we propose minimum regret search (MRS), a novel acquisition
function that explicitly aims at minimizing the expected simple regret
(Section \ref{Section:MRS}). MRS performs well according to both the mean and
median performance on a synthetic problem (Section
\ref{Section:ResultsSingleTask}). Moreover, we discuss how MRS can be extended
to multi-task optimization problems (Section \ref{Section:CMRS}) and present
empirical results on a simulated robotic control problem (Section
\ref{Section:ResultsMultiTask}).

@@

While information-based policies for Bayesian optimization such as ES and PES
have performed very well empirically, their internal objective of minimizing
the uncertainty in the location of the optimizer $\mathbf{x}^\star$, i.e.,
minimizing the differential entropy of $p_\star$, is actually different (albeit
related) to the common external objective of minimizing the simple regret of
$\mathbf{\tilde x}_N$, the recommendation of BO for $\mathbf{x}^\star$ after $N$
trials, where we define the simple regret of $\mathbf{\tilde x}_N$ as
$R_f(\mathbf{\tilde  x}_N) = f(\mathbf{x}^\star) - f(\mathbf{\tilde  x}_N) = \max_{\mathbf{x}} f(\mathbf{x}) - f(\mathbf{\tilde x}_N)$.

Clearly, $\mathbf{x}^\star$ has zero and thus minimum simple regret, but a query that
results in the maximal decrease in $H(\mathbf{x}^\star)$ is not necessarily the
one that also results in the maximal decrease in the expected simple regret
as we will discuss in Section \ref{Section:MRS_Illustration}. In this section, we propose \emph{minimum
regret search} (MRS), which explicitly aims at minimizing the expected simple
regret.

\subsection{Formulation}
Let $\mathcal{X} \subset \mathcal{R}^D$ be some bounded domain and $f: \mathcal{X} \mapsto
\mathbb{R}$ be a function. We are interested in finding the maximum
$\mathbf{x}^\star$ of $f$ on $\mathcal{X}$. Let there be a probability measure
$p(f)$ over the space of functions $f: \mathcal{X} \mapsto \mathbb{R}$, such as
a GP. Based on this $p(f)$, we would ultimately like to
select an $\mathbf{\tilde x}$ which has minimum simple regret
$R_f(\mathbf{\tilde x})$. We define the expected simple regret ER of
selecting parameter $\mathbf{x}$ under $p(f)$ as follows:
$$\text{ER}(p)(\mathbf{x}) = \mathbb{E}_{p(f)}[R_f(\mathbf{x})]
= \mathbb{E}_{p(f)}[\max_{\mathbf{x}} f(\mathbf{x}) - f(\mathbf{
x})]$$
The minimizer $\mathbf{\tilde x}$ of the expected simple regret under fixed $p(f)$ can be determined efficiently in the case of a GP since it is identical to the maximizer of the GP's mean $\mu_N(\mathbf{x})$. However, in general, $p(f)$ depends on data and it is desirable to select the data such that $\text{ER}$ is also minimized with regard to the resulting $p(f)$.

@@

As choosing the optimal sequence $X^q_N$ at once is intractable, we follow the
common approach in BO and select $\mathbf{x}^q_n$
sequentially in a myopic way. However, as $\mathbf{\tilde x}_N$ itself depends
on $\mathcal{D}_N$ and is thus unknown for $n < N$, we have to use proxies for it based on the currently
available subset $\mathcal{D}_n \subset \mathcal{D}_N$.  One simple choice for a proxy is to use
the point which has minimal expected simple regret under $p_n(f) =
p(f \vert \mathcal{D}_n)$, i.e.,
$\mathbf{\tilde x} = \arg\min_\mathbf{x} \text{ER}(p_n)(\mathbf{x})$.
%Based on this, we define the one-step
%(myopic) expected regret reduction with respect to a given $p(f)$ when
%evaluating the function at $\mathbf{x}^q$.
Let us denote the updated
probability measure on $f$ after performing a query at $\mathbf{x}^q$ and
observing the function value $y=f(\mathbf{x}^q) + \epsilon$ by
$p^{[\mathbf{x}^q, y]}_n = p(f\vert \mathcal{D}_n \cup \{\mathbf{x}^q, y\})$.
We define the acquisition function MRS$^{\text{point}}$ as the expected reduction of the minimum expected regret for a query at
$\mathbf{x}^q$, i.e.,
\begin{equation*}
\begin{split}
a_{\text{MRS}^{\text{point}}}(\mathbf{x}^q)
    & = \min_{\mathbf{\tilde x}}\text{ER}(p_n)(\mathbf{\tilde x}) \\
     &- \mathbb{E}_{y \vert p_n(f), \mathbf{x}^q}[\min_{\mathbf{\tilde x}}  \text{ER}(p^{[\mathbf{x}^q, y]}_n)(\mathbf{\tilde x})],
\end{split}
\end{equation*}
where the expectation is with
respect to $p_n(f)$'s predictive distribution at $\mathbf{x}^q$ and we drop the
implicit dependence on $p_n(f)$ in the notation of $a_{\text{MRS}^{\text{point}}}$.
The next query point $\mathbf{x}^q_{n+1}$ would thus be selected as the maximizer of
$a_{\text{MRS}^{\text{point}}}$.

@@

MRS can thus be seen as a more Bayesian treatment, where we marginalize our
uncertainty about $\mathbf{\tilde x}_N$, while MRS$^\text{points}$ is more akin
to a point-estimate since we use a single point (the minimizer of the expected simple regret)
as proxy for $\mathbf{\tilde x}_N$.

@@

Figure \ref{fig:MRS_illustration} presents an illustration of different
acquisition functions on a simple one-dimensional target function. The left
graphic shows a hypothetical GP posterior (illustrated by its mean and standard
deviation) for length scale $l=0.75$, and the resulting probability of $\mathbf{x}$
being the optimum of $f$ denoted by $p^\star$.
Moreover, the expected simple regret of selecting $\mathbf{x}$ denoted by
$\text{ER}(\mathbf{x})$ is shown. The minimum of $\text{ER}(\mathbf{x})$ and the maximum
of $p^\star$ are both at $\mathbf{\tilde x}=1.5$. The expected regret of $\mathbf{\tilde x}$
is approximately $\text{ER}(\mathbf{\tilde x}) = 0.07$. We plot $\mathbb{E}[\max(f(\mathbf{x}) -
f(\mathbf{\tilde x}), 0)]$ additionally to shed some light onto situations in which $
\mathbf{\tilde x} = 1.5$ would incur a significant regret: this quantity shows that most of the
expected regret of $\mathbf{\tilde x}$ stems from situations where the ``true''
optimum is located at $\mathbf{x} \geq 3.5$. This can be explained by the
observation that this area has high uncertainty and is at the same time largely
uncorrelated with $\mathbf{\tilde x} = 1.5$ because of the small length-scale of the GP.

@@

\begin{figure*}
\centering
\includegraphics[width=.9\textwidth]{empirical_comparison}
\includegraphics[width=.9\textwidth]{hist}
\caption{(Top) Median and mean simple regret over $250$ repetitions for different acquisition functions. Shown is the simple regret of the recommendation $\mathbf{\tilde x}_N$ after $N$ trials, i.e., the point which maximizes the GP posterior
mean. (Bottom) Histogram of the simple regret after performing $N=100$ trials for different acquisition functions (note the log-scales).}
\label{fig:empirical_comparison}
\end{figure*}

@@

Figure \ref{fig:empirical_comparison} summarizes the results for a pure
exploration setting, where we are only interested in the quality of the
algorithm's recommendation for the optimum after $N$ queries but not in the
quality of the queries themselves:  according to the
median of the simple regret (top left), ES, MRS, and MRS$^{point}$ perform
nearly identical, while EI is about an order of magnitude worse. GP-UCB performs
even worse initially but surpasses EI eventually\footnote{GP-UCB reaches 
the same level of mean and median simple regret  as MRS after $N=200$ steps 
(in mean and median) with no significant difference  according to a Wilcoxon signed-rank test.}. 
PI performs the worst as it
exploits too aggressively. These results are roughly in correspondence with prior
results \cite{hennig_entropy_2012,hernandez-lobato_predictive_2014} on the same
task; note, however, that \citet{hernandez-lobato_predictive_2014} used a lower
noise level and thus, absolute values are not comparable. However, according to
the \emph{mean} simple regret, the picture changes considerably (top right):
here, MRS, MRS$^{point}$, and EI perform roughly on par while ES is about an order of
magnitude worse. This can be explained by the distribution of the simple
regrets (bottom): while the distributions are fairly non-normal for all
acquisition functions, there are considerably more runs with very high simple
regret ($R_f(\mathbf{\tilde x}_N)>10^{-2}$) for ES (10) than for MRS (4) or EI (4).

We illustrate one such case where ES incurs high simple regret in Figure
\ref{fig:es_analysis}. The same set of representer points has been used for ES and MRS.
While both ES and MRS assign a non-zero density $p^\star$
(representer points) to the area of the true optimum of the function
(bottom center), ES assigns high acquisition value only to areas with a high
density of $p^\star$ in order to further concentrate density in these areas.
Note that this is not due to discretization of $p^\star$ but because of ES'
objective, which is to learn about the precise location of the optimum, irrespective of
how much correlation their is between the representer points according to $p(f)$.
Thus, predictive entropy search \cite{hernandez-lobato_predictive_2014} would likely be affected by the
same deficiency. In contrast, MRS focuses first on areas which have not been
explored and have a non-zero $p^\star$, since those are areas with high expected
simple regret (see Section \ref{Section:MRS_Illustration}). Accordingly, MRS is less likely to incur a high simple regret. In summary, the MRS-based acquisition functions are the only
acquisition functions that perform well both according to the median and the
mean simple regret; moreover, MRS performs slightly better than
MRS$^{point}$ and we will focus on MRS subsequently.
